---
title: "Coursera - Practical Machine Learning - Final Project"
author: "Daniel Morales"
date: "2020-08-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Using devices such as *Jawbone Up*, *Nike FuelBand*, and *Fitbit* it is now 
possible to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement – 
a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how *much* of a 
particular activity they do, but they rarely quantify *how well they do it*. 

In this project, our goal will be to use data from accelerometers on the belt, 
forearm, arm, and dumbbell of 6 participants. They were asked to perform barbell 
lifts correctly and incorrectly in 5 different ways: exactly according to the 
specification (class A), throwing the elbows to the front (class B), lifting the 
dumbbell only halfway (class C), lowering the dumbbell only halfway (class D) 
and throwing the hips to the front (class E). More information is available from 
the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on 
the Weight Lifting Exercise Dataset).

This data will be used to train a model that predicts which way (A, B, C, D or 
E) the exercise is being done, given the measurements.

## Data

The data used a courtesy of:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative 
Activity Recognition of Weight Lifting Exercises. Proceedings of 4th 
International Conference in Cooperation with SIGCHI (Augmented Human '13). 
Stuttgart, Germany: ACM SIGCHI, 2013.

The labeled data for this project are available here (this is the data provided
for us to train and test our model in this project):

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The unlabeled data are available here (and the main goal of the project is to
correctly label these observations with our trained model):

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

### Loading Libraries and Importing Data

We start loading packages and reading data directly from the links provided. 
Note that the first row contains headers and the first column contains row 
numbers.

```{r packages, message = FALSE, results = "hide", cache = TRUE}
library(caret)
library(corrplot)
library(e1071)
library(rattle)
```

```{r import, cache = TRUE}
labeled <- read.csv(
  url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"),
  header = TRUE, row.names = 1
)
unlabeled <- read.csv(
  url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"),
  header = TRUE, row.names = 1
)
dim(labeled)
dim(unlabeled)
```

### Cleaning Data

After importing the datasets with `r dim(unlabeled)[2]+1` columns each 
(`r dim(unlabeled)[2]` if we do not count the column with row numbers), matching 
every column except for the last one, which is `classe`, our target variable 
for the labeled set, and ```problem_id``` for the unlabeled set, an ID for 
submitting the predictions. We also have `r dim(labeled)[1]` rows 
(observations) for the labeled set and `r dim(unlabeled)[1]` rows for the 
unlabeled set.

#### Missing Values

A simple `str` applied to the data (due to number of variables, this step is not 
shown) shows us that there are variables (columns) with most values as `NA` or empty `""`. These variables will be disconsidered.

```{r clean_missing, cache = TRUE}
labeled_clean <- labeled[, colSums(is.na(labeled) | labeled == "") < 0.95 * nrow(labeled)]
unlabeled_clean <- unlabeled[, colSums(is.na(unlabeled)) < 0.95 * nrow(unlabeled)]
dim(labeled_clean)
dim(unlabeled_clean)
```

#### Identification Variables

The first 6 columns will also be disconsidered, as they are identification 
variables, and our goal is to predict based entirely on the movements captured 
by the accelerometers.

```{r clean_id_var, cache = TRUE}
labeled_clean <- labeled_clean[, -c(1:6)]
unlabeled_clean <- unlabeled_clean[, -c(1:6)]
dim(labeled_clean)
dim(unlabeled_clean)
```

#### High Correlations

Let us now analyze the correlations between the observations.

```{r correlation, cache = TRUE}
corrplot(cor(labeled_clean[, -53]), order = "FPC", method = "color",
         type = "upper", tl.cex = 0.5, tl.col = rgb(0, 0, 0))
```

Being the correlations quite few, we will refrain from doing a more profound 
analysis on the correlations, like PCA (Principal Component Analysis) and just 
proceed to the next steps.

### Partition Data

We divide our labeled data in training and testing sets at the ratio of 80% for 
training and 20% for testing, followed by a 5-fold cross validation. Both values 
have empirically shown performance that balance well bias and variance. 

```{r partition, cache = TRUE}
set.seed(8226)
in_train <- createDataPartition(labeled_clean$classe, p = 0.8, list = FALSE)
training <- labeled_clean[in_train, ]
testing <- labeled_clean[-in_train, ]
tr_control <- trainControl(method = "cv", number = 5)
```

## Prediction Algorithms

We will be using three different algorithms for our multiclass classification:
a simple classification tree for reference, random forest and GBM (Gradient 
Boosting Machine).

### Classification Tree

```{r classif_tree, cache = TRUE}
rpart_fit <- train(classe ~ ., data = training, 
                   method = "rpart", trControl = tr_control)
plot(rpart_fit)
fancyRpartPlot(rpart_fit$finalModel, sub = NULL)
rpart_predict <- predict(rpart_fit, newdata = testing)
rpart_conf_mat <- confusionMatrix(rpart_predict, factor(testing$classe))
rpart_conf_mat$table
```

Note that this simple classification tree never predicts class D!

### Random Forest

```{r random_forest, cache = TRUE}
rf_fit <- train(classe ~ ., data = training, 
                method = "rf", trControl = tr_control)
plot(rf_fit)
rf_predict <- predict(rf_fit, newdata = testing)
rf_conf_mat <- confusionMatrix(rf_predict, factor(testing$classe))
rf_conf_mat$table
```

### Gradient Boosting Machine

```{r gbm, cache = TRUE}
gbm_fit <- train(classe ~ ., data = training, 
                method = "gbm", trControl = tr_control, verbose = FALSE)
plot(gbm_fit)
gbm_predict <- predict(gbm_fit, newdata = testing)
gbm_conf_mat <- confusionMatrix(gbm_predict, factor(testing$classe))
gbm_conf_mat$table
```

## Prediction

Since we have quite balanced classes on our target variable `classe`, we can use 
accuracy on the testing set as a measure to select a prediction algorithm. 
Having:

| Algorithm           | Accuracy                                 |
|---------------------|------------------------------------------|
| Classification Tree | `r rpart_conf_mat$overall[["Accuracy"]]` |
| Random Forest       | `r rf_conf_mat$overall[["Accuracy"]]`    |
| GBM                 | `r gbm_conf_mat$overall[["Accuracy"]]`   |

So the selected algorithm for prediction is random forest, that gives the 
following predictions for our unlabeled dataset:

```{r prediction_rf, cache = TRUE}
predict(rf_fit, newdata = unlabeled)
```

Note that the second best algorithm (according to accuracy), GBM, produces the 
same predictions:

```{r prediction_gbm, cache = TRUE}
predict(gbm_fit, newdata = unlabeled)
```

Due to the high accuracy from the random forests we can keep ourselves within
these somewhat simple algorithms and within the scope of the course.